Last glacial cycle glacier erosion potential in the Alps

### 2021-02-04 articles/erosion/alps-erosion.tex
Rework intro and integrate Ian's comments.

\documentclass[esurf, manuscript]{copernicus}

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, mountaineers and scientists for centuries. Its cultural
    impact is indeed so far-reaching, that in English, a non-Alpine language,
    the adjective ``alpine'' with non-capital ``a'' is now casually used to
    describe an Alpine-like, glacially modified mountain landscape outside the
    Alps, while the proper noun ``Alps'' has been applied to nick-name
    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
    New Zealand (Southern Alps), Japan
    (\emph{%\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
    Nihon Arupusu}), and elsewhere.

    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps, present a higher variety glacial erosional landforms
    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
    be understood.
    In absence of other glacial landforms, the degree of glacial modification
    of a landscape has sometimes been used as a proxy for the duration of past
    glacier cover, as for instance in geomorphological mapping
    \citep[e.g.][]{Margold.etal.2011, Fu.etal.2012}. Cold-based glaciers have
    been shown to preserve landforms as fragile as sand beaches
    \citep{Kleman.1994}. Yet it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    glacially preserved landscapes and others, including the Alps, do not
    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

    \ian{Furthermore, glacier erosion likely  governs, at least in part, the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through the ``glacier buzz-saw''.
      In turn, evidence suggests that increase erosion rates coincided with global cooling and  Pliestocene glaciations from roughly 2.6\unit{Ma} Bp \citep{Herman.Champagnac.2016}.
      Yet, other work suggests that global weathering rates remained constant over this time period \citep{Willenbring.Von-Blanckenburg.2010}.
      Examination of contemporary erosion rates across different climates suggest that increased temperatures in glaciated regions lead to higher erosion rates and that erosion rates may have increased towards present \citep{Koppes.Montgomery.2009,Koppes.etal.2015,Fernandez.etal.2016}.
      However, understanding the link between climate and glacier erosion in the sedimentary record remains difficult due to time scale biases \citep{Ganti.etal.2016} and the non-linear relationship  between temperature and erosion \citep[e.g.][]{Anderson.etal.2012,Mariotti.etal.2021}}

    \ian{
      Observing and quantifying the long-term erosion
      and sedimentation processes has been a challenge, and two general methods have been adopted to quantify erosion \citep{Alley.etal.2019}.
      Physically-based models describe quarrying process and abrasion of bedrock by debris laden ice \citep[e.g.][]{Alley.etal.1997,Iverson.2012,Beaud.etal.2014}.
      Describing the processes physically provides a basis for understanding erosion processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into the formation of glacial landforms, such as tunnel valleys and eskers \citep{Beaud.etal.2018,Hewitt.Creyts.2019}.
      Yet, these models prove difficult to implement in many cases due to the large number of poorly constrained parameters and process, i.e. water pressure flucuations and subglacial debris concentration \citep[e.g.][]{Hallet.1979,Ugelvig.etal.2018}.}

    Instead, empirical relationships between glacier sliding and erosion can represent erosional quantities  well \citep[e.g.][]{Humphrey.Raymond.1994, Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial landforms \citep[e.g.][]{Harbor.etal.1988,Macgregor.etal.2000}.
    \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a near-square relationship between basal sliding and glacier
    erosion, but yielding much higher erosion rates.
    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear relationship to erosion.

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. Despite aggregated uncertainties on paleoclimate,
    glacier flow and glacier erosion processes, out results provide insights
    into the diversity of the Alpine glacial erosion landscape. \ian{Furthermore,
      we examine the glaciologcial conditions that cause erosion and discuss
      the implications of these conditions on understanding the relationship between climate and glacier erosion.  }

>>>>

\documentclass[esurf, manuscript]{copernicus}
\usepackage{CJK}  % Japanese text

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, mountaineers and scientists for centuries. Its cultural
    impact is indeed so far-reaching, that in English, a non-Alpine language,
    the adjective ``alpine'' with non-capital ``a'' is now casually used to
    describe an Alpine-like, glacially modified mountain landscape outside the
    Alps, while the proper noun ``Alps'' has been applied to nick-name
    Alpine-like, glacier eroded mountain ranges in Norway (\emph{Lyngsalpene}),
    New Zealand (Southern Alps), Japan
    (\emph{\begin{CJK}{UTF8}{min}日本アルプス\end{CJK},
    Nihon Arupusu}), and elsewhere.

    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps, present a higher variety glacial erosional landforms
    (Fig.~\ref{fig:landscape}), whose implications on glacial history is yet to
    be understood.

    In absence of other landforms, glacially eroded topography has sometimes
    been used as a proxy for glacier extension, for instance in
    geomorphological mapping \citep[e.g.,][]{Margold.etal.2011, Fu.etal.2012}.
    Cold-based glaciers, however, have been observed to preserve landforms as
    fragile as sand beaches \citep{Kleman.1994}, leading to the
    reinterpretation of complex glacial landscapes within a binary conceptual
    framework of cold and temperate basal thermal regimes
    \citep{Kleman.etal.2008, Kleman.etal.2010, Fabel.etal.2012, Fu.etal.2013}.
    Nevertheless, it remains unclear how sharp the erosional
    transition to cold-based ice is, and why some glaciated regions show
    glacially preserved landscapes and others, including the Alps, do not
    \citep{Wirsig.etal.2016, Seguinot.etal.2018}.

    On longer time scales, glacier erosion likely governs, at least in part,
    the height of mountains \citep{Egholm.etal.2009,Thomson.etal.2010} through
    the ``glacier buzz-saw''. In turn, evidence suggests that increased
    mountain erosion rates coincided with global cooling and Pliestocene
    glaciations from roughly 2.6\,Ma (million years before present)
    \citep{Herman.Champagnac.2016}. Yet, other work suggests that global
    weathering rates remained constant over this time period
    \citep{Willenbring.Von-Blanckenburg.2010}. \note{Evidence suggests, other
    work suggests, maybe this can be reworked.} Examination of contemporary
    erosion rates across different climates suggest that increased temperatures
    in glaciated regions lead to higher erosion rates and that erosion rates
    may have increased towards present \citep{Koppes.Montgomery.2009,
    Koppes.etal.2015, Fernandez.etal.2016}. However, understanding the link
    between climate and glacier erosion in the sedimentary record remains
    difficult due to time scale biases \citep{Ganti.etal.2016} and the
    non-linear relationship  between temperature and erosion
    \citep[e.g.,][]{Anderson.etal.2012,Mariotti.etal.2021}.

    Observing and quantifying the long-term glacier erosion and sedimentation
    processes has been a challenge, and two general methods have been adopted
    to quantify erosion \citep{Alley.etal.2019}. Physically-based models
    describe quarrying process and abrasion of bedrock by debris laden ice
    \citep[e.g.,][]{Alley.etal.1997, Iverson.2012, Beaud.etal.2014}. Describing
    the processes physically provides a basis for understanding erosion
    processes \citep{Hallet.1979, Ugelvig.etal.2018} and yields insight into
    the formation of glacial landforms, such as tunnel valleys and eskers
    \citep{Beaud.etal.2018, Hewitt.Creyts.2019}. Yet, these models prove
    difficult to implement in many cases due to the large number of poorly
    constrained parameters and process, i.e. water pressure flucuations and
    subglacial debris concentration \citep[e.g.,][]{Hallet.1979, Seguinot.2008,
    Ugelvig.etal.2018}.

    Instead, empirical relationships between glacier sliding and erosion can
    represent erosional quantities well \citep[e.g.,][]{Humphrey.Raymond.1994,
    Herman.etal.2015, Koppes.etal.2015} and can recreate important glacial
    landforms \citep[e.g.,][]{Harbor.etal.1988, Macgregor.etal.2000}.
    \citet{Koppes.etal.2015} quantified sediment
    yields in 15~Patagonian and Antarctic Peninsula fjords, concluding at a
    predominant control of surface air temperature on rapid glacier erosion and
    calibrating a near-square-law from glacier velocity to erosion rate.
    \citet{Herman.etal.2015} collected suspended sediment samples for 5~months
    in the outlet stream of Franz-Josef Glacier, mapped their chemical
    composition to geologic zones of distinctive glacier speed, and also
    concluded to a near-square relationship between basal sliding and glacier
    erosion, but yielding much higher erosion rates.
    \citet{Cook.etal.2020} assembled a global compilation of erosion rates for
    38~glaciers, showing a predominant role of glacier sliding velocity over
    climate variables, yet concluding at a sub-linear relationship to erosion.

    Here, the non-linear erosion law by \citet{Koppes.etal.2015} is applied to
    previously published model results \citep{Seguinot.etal.2018} and the
    patterns of modelled erosion rate and cumulative last glacial cycle erosion
    potential are analysed. We examine the glaciologcial conditions that cause
    erosion and discuss the implications of these conditions on understanding
    the relationship between climate and glacier erosion, Despite aggregated
    uncertainties on paleoclimate, glacier flow and glacier erosion processes,
    out results provide insights into the diversity of the Alpine glacial
    erosion landscape.
